---
title: "Reproducible Research Project 1"
output: html_document
---

Required packages
plyr


Loading and preprocessing the data:
```{r}
unzip("./repdata-data-activity.zip")
activity<-read.csv("./activity.csv")
```


What is the total number of steps taken per day?
```{r}
library(plyr)
step_sum_mean<-ddply(activity,~date, summarize, sum=sum(steps, na.rm=FALSE), mean=mean(steps, na.rm=TRUE))
step_sum<-sum(step_sum_mean$sum, na.rm=TRUE)
step_mean<-mean(step_sum_mean$mean, na.rm=TRUE)
```

The total number of steps taken per day (sum) and the mean steps are listed below:
```{r, echo=FALSE}
step_sum_mean
```

The total number of steps taken on all days is listed below:
```{r, echo=FALSE}
step_sum
```


The mean number of steps taken across all days is listed below:
```{r, echo=FALSE}
step_mean
```

Histogram of the total number of steps per day:
```{r, echo=FALSE}
hist(step_sum_mean$sum, breaks=20, col="red", main= "Distribution of Total Number of Steps per Day",
     xlab="Total Number of Steps per Day",
     ylab="Frequency (Days)")
```

Calculate the mean and median of the total number of steps per day:
```{r}
mean_median_activity<-ddply(activity, ~date, summarize, mean=mean(steps, na.rm=TRUE), median=median(steps,na.rm=TRUE))
median_activity<-median(activity$steps, na.rm=TRUE)
```

The mean and median steps per day are listed below:
```{r, echo=FALSE}
mean_median_activity
```

The median steps across all days are listed below:
```{r, echo=FALSE}
median_activity
```

The mean number of steps taken across all days is listed below:
```{r, echo=FALSE}
step_mean
```

What is the average daily activity pattern?

Time series plot of the 5-minute interval and average number of steps taken, averaged across all days
```{r}
step_sum_by_interval<-ddply(activity,~interval, summarize, mean=mean(steps,na.rm=TRUE))
```
```{r, echo=FALSE}
plot(step_sum_by_interval$interval, step_sum_by_interval$mean, type="l",
     xlab="5 minute interval", ylab="Mean Number of Steps")
```

Activity appears to peak between intervals 600 and 1000 based upon the time series plot.

Which 5-min interval, on average across all days, contains the maximum number of steps?
```{r}
max_interval<-which.max(step_sum_by_interval$mean)
```

The 5-min interval that contains the maximum number of steps is listed below:
```{r, echo=FALSE}
step_sum_by_interval[max_interval,1]
```

Inputting missing values

Calculate and report the total number of missing values in the dataset.
```{r}
NAsum<-sum(length(which(is.na(activity$steps))))
```

The total number of missing values in the dataset is listed below:
```{r, echo=FALSE}
NAsum
```


Fill in all missing values (coded as NA) in the dataset yielding a new, equal dataset called NAreplace_activity.
```{r}
NAreplace_activity<-activity
for (i in 1:nrow(NAreplace_activity)) {
  if (is.na(NAreplace_activity$steps[i])) {
    NAreplace_activity$steps[i]<-step_sum_by_interval[which(NAreplace_activity$interval[i]==step_sum_by_interval$interval), ]$mean
  }
}
```

Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day in NAreplace_activity.  
```{r, echo=FALSE}
hist(step_sum_mean$sum, breaks=20, col="red", main= "Distribution of Total Number of Steps",
     xlab="Total Number of Steps in a Day",
     ylab="Frequency in Days")
```


Calculate and report the mean and median total number of steps taken per day in the new data set in which the NAs have been replaced.

```{r}
mean_median_NAreplace<-ddply(NAreplace_activity, ~date, summarize, mean=mean(steps), median=median(steps))
meanNAreplace<-mean(NAreplace_activity$steps)
medianNAreplace<-median(NAreplace_activity$steps)
sumStepsNAreplace<-sum(NAreplace_activity$steps)
sumStepsNAreplace_daily<-ddply(NAreplace_activity, ~date, summarize, NAreplace_steps=sum(steps))
```

The mean and median of the new data set in which the NAs have been replaced are listed below:
```{r, echo=FALSE}
mean_median_NAreplace
```

The mean across all days is listed below:
```{r}
meanNAreplace
```

The median across all days is listed below:
```{r}
medianNAreplace
```


Do these values differ from the estimates from the first part of the assignment?
Answer: These values do not appear to be different.

What is the impact of imputing missing data on the estimates of the total daily number of steps?

The total number of steps per day in the original activity data compared to the step sums in the new data set with NAs removed is listed below:
```{r, echo=FALSE}
cbind(step_sum_mean$date, step_sum_mean$sum, sumStepsNAreplace_daily$NAreplace_steps) 
```

Are there differences in activity patterns between weekdays and weekends?

Answer: The only differences seem to exist where NAs were previously present in the original dataset, otherwise the values appear to be identical

Create a new factor variable in the dataset with two levels-"weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activity$date<-as.POSIXct(activity$date)
activity$WeekDay<-weekdays(activity$date)
activity$DayFactor<-ifelse(activity$WeekDay%in%c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "weekday", "weekend")
```

Make time-series plots of the 5-minute activity intervals and average number of steps taken to compare weekday and weekend activity. 
```{r}
weekday_activity<-activity[activity$DayFactor=="weekday",]
weekday_activity<-ddply(weekday_activity, ~interval, summarize, mean=mean(steps, na.rm=TRUE))

weekend_activity<-activity[activity$DayFactor=="weekend",]
weekend_activity<-ddply(weekend_activity, ~interval, summarize, mean=mean(steps, na.rm=TRUE))
```

```{r, echo=FALSE}
par(mfrow=c(2,1))
plot(weekend_activity$interval, weekend_activity$mean, type="l", main="Weekend Activity", xlab="5 min Interval ", ylab="Mean Steps")
plot(weekday_activity$interval, weekday_activity$mean, type="l", main="Weekday Activity", xlab="5 min Interval", ylab="Mean Steps")
```

Activity peaks between intervals 600 and 1000 but weekend activity remains high throughout the day.